---
title: "phosphocie: creating kinase colour representations from phosphosites"
author: "Cas Blaauw"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{phosphocie}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 0. Overview
`phosphocie` is a package to create kinase colour embeddings: continuous colours that 
denote kinase specificities for a phosphorylation site. These colour embeddings are
based on the [U-CIE framework](https://doi.org/10.1101/2021.12.02.470966):
reduce data down to 3 dimensions and optimally embed it in the CIELAB colourspace.
Specifically, `phosphocie` provides functions to handle phosphorylation site data,
create kinase prediction tool input files, handle the resulting kinase prediction output,
and a pre-trained kinase colourspace based on the full 
[PhosphoSitePlus](https://www.phosphosite.org/) phosphorylation site dataset

In this vignette, we will demonstrate `phosphocie` using the kinase-substrate-annotated 
phosphorylation site dataset, from [PhosphoSitePlus](https://www.phosphosite.org/). 

```{r setup}
library(phosphocie)
library(dplyr)
library(ggplot2)
library(umap)
```

## 1. Prepare phosphorylation site data
### Read in phosphorylation site data


The base data can be loaded in with `read_kinsub()` and filtered to valid human sequences
with `filter_phosphosites()`:

```{r}
kinsub_path <- system.file('extdata', 'kinase_substrate_dataset_head', package = 'phosphocie')
kinsub <- read_kinsub(kinsub_path)

head(kinsub_human)
```

### Build FASTAs
`build_fastas()` creates a FASTA file with a separate FASTA entry (header and sequence)
for each site, which can be used for input in NetPhorest. 

```{r}
tmp <- tempdir()
build_fastas(kinsub_human, name_col = 'unique_id', seq_col = 'fragment_15', path = file.path(tmp, 'kinsub_human_peptides.fasta'))
```

Alternatively, `retrieve_fastas()` retrieves full fasta files for a list of accession IDs.
However, `phosphocie` is geared towards visualising kinase predictions for individual, 
possibly unidentified sites, so this option is not recommended and requires more manual processing
to work with the follow-up functions.

```{r, eval = FALSE}
retrieve_fastas(kinsub_human$acc_id, file.path(tmp, 'kinsub_human.fasta'))
```

## Predict kinases

Kinase prediction can be done with your tool of choice, but phosphocie natively supports
NetPhorest. For details on how to run NetPhorest locally, see [the NetPhorest site](http://netphorest.info/download.shtml).

```{}
# Run NetPhorest
cat kinsub_human.fasta | ./netphorest > kinsub_human_netphorest
```

This output file is also included in the package, and can be read in using 
`read_netphorest()` and `filter_netphorest()`:
```{r}
kinsub_netphorest_path <- system.file('extdata', 'kinsub_head_netphorest', package = 'phosphocie')
kinsub_netphorest_raw <- read_netphorest(kinsub_netphorest_path) 

kinsub_netphorest <- filter_netphorest(kinsub_netphorest_raw, source_window_size = 15)
```

`filter_netphorest()` attempts to filter out sites in our sequences from `build_fasta()` 
that were detected by NetPhorest, but that aren't the true intended sites in the middle
of the supplied sequence window. 
It can determine conclusive sites for any non-truncated windows or windows truncated by the 
end of the protein, but struggles to distinguish between true sites at the start of a protein
and false sites at the beginning of the window without additional data.

For example, we can take a look at one uncertain ID in this data:
```{r}
uncertain <- kinsub_netphorest_raw %>% filter(fasta_id == 'P13796|LCP1|L-plastin|S5|ARGSVSD')
uncertain
```

We know from the metadata encoded in `fasta_id` that S5 is the true site.
It's not at position 8, where we normally expect sites in data with a `source_window_size`
of 15 like PhosphoSitePlus, because the original 'filler' was truncated.
However, without this metadata knowledge, it's impossible to distinguish this site
from a fake detected site at the start of a window, in which case position 7 would 
be the better option.

By default, `filter_netphorest()` will use the included site fragment included in the
fasta header by `build_fastas()` to detect the true site:

```{r}
uncertain %>%
  tidyr::extract(fasta_id, c('orig_res', 'orig_site'), '\\|(\\w)(\\d{1,4})$', remove = FALSE, convert = TRUE) %>% 
  filter_netphorest(source_window_size = 15, protein_res_col = 'orig_res', protein_pos_col = 'orig_site')

uncertain %>%
  filter_netphorest(source_window_size = 15)
```

This can also be passed as a manual column:

```{r}
uncertain %>%
  tidyr::extract(fasta_id, 'fragment', '\\|([A-Za-z_-])$', remove = FALSE) %>% 
  filter_netphorest(source_window_size = 15, fragment_col = 'fragment')
```


If this metadata is not available, `keep_uncertain` determines how remaining uncertainties are handled:
* `keep_uncertain = NULL` (default): Keep the site closest to the middle.
* `keep_uncertain = TRUE`: Don't filter uncertain sites, and keep all possibilities instead.
* `keep_uncertain = FALSE`: Drop any uncertain sites completely, only return sites with certainty.

```{r}
uncertain %>% filter_netphorest(source_window_size = 15, keep_uncertain = NULL)
uncertain %>% filter_netphorest(source_window_size = 15, keep_uncertain = TRUE)
uncertain %>% filter_netphorest(source_window_size = 15, keep_uncertain = FALSE)
```


## Create colour embeddings

### Join together the original data and kinase predictions

We match by the ID column, which is the primary identifier carried over from original data
to the prediction file. 

However, if these are distinct but both contain information that can
be used to match (like gene/protein name, accession id, original protein location/residue), 
this information can be extracted using a function like `tidyr::separate()` and the resulting
columns can be used to match as well (i.e. `by = c('acc_id', 'position', 'residue')`)

```{r}
kinsub_full <- left_join(kinsub_human, kinsub_netphorest, by = c('unique_id' = 'fasta_id')) 
kinsub_missing <- kinsub_full %>% filter(is.na(fragment_11)) # Data for which no prediction could be recovered
kinsub_full <- kinsub_full %>% filter(!is.na(fragment_11))
```

### Separate data into labels and predictions

```{r}
kinsub_preds <- select(kinsub_full, unique_id, Abl_group:YSK_group) %>% 
  tibble::column_to_rownames('unique_id') %>% 
  as.matrix()
kinsub_info <- select(kinsub_full, -c(Abl_group:YSK_group))
```

### Transform data using the reference U-CIE space

This package includes a pre-calculated kinase colour embedding space, based on 
NetPhorest kinase predictions for all sites in [PhosphoSitePlus](https://www.phosphosite.org/)'s
phosphorylation site dataset.

```{r}
# Transform with reference UMAP
kinsub_preds_3d <- predict(ref_umap, kinsub_preds)

# Transform with reference U-CIE
kinsub_ucie <- kinase2cielab(kinsub_preds_3d, ref_transform)$hex
```

### Alternative: transform into standalone U-CIE space

We can also fit your dataset to a new colourspace, based solely on (the 3D representation of) the dataset.
```{r, eval = FALSE}
umap_settings_3d <- umap::umap.defaults
umap_settings_3d$n_components <- 3
kinsub_umap_3d <- umap::umap(kinsub_preds, umap_settings_3d)

kinsub_transformations <- ucie_transformations(kinsub_umap_3d$layout)
kinsub_ucie <- kinase2cielab(kinsub_umap_3d$layout, kinsub_transformations)$hex
```


## Visualise kinase colour embeddings

### Reduce data to 2D
```{r}
kinsub_umap_2d <- umap(kinsub_preds, umap.defaults)
kinsub_2d <- kinsub_umap_2d$layout %>% magrittr::set_colnames(c('X', 'Y')) 
kinsub_plot_data <- bind_cols(kinsub_info, kinsub_2d, ucie = kinsub_ucie)
```

### Create plots
```{r}
kinsub_plot_data %>% 
  slice_sample(n = 1000) %>% 
  ggplot(aes(X, Y, color = ucie)) + 
  geom_point(size = 3) + 
  scale_color_identity() + 
  guides(color = 'none')

kinsub_plot_data %>% 
  mutate(kin_gene = forcats::fct_lump_n(kin_gene, 15)) %>% 
  ggplot(aes(X, Y, color = ucie)) + 
  geom_point(size = 3) + 
  scale_color_identity() + 
  guides(color = 'none') + 
  facet_wrap(~kin_gene)
```

